---
title: "impala.revision.noran"
author: "Matt Wuensch"
date: "2025-05-07"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
set.seed(63)
library(dplyr)
library(lme4)
library(lmerTest)
library(emmeans)
```
# Analyses for 2017, comparing differences in foraging and vigilance for males, females, and juveniles in territorial herds in WRF

```{r}
# Bring in 2017 data
imp.17 <- read.csv("2017_individual.csv")

# Making new column of non-foraging and non-vigilance. needed for binomial regressions wherein we have success/failure format meaning success = targeted behavior and failure = not-targeted behavior.
imp.17$non.forage <- imp.17$vigilant + imp.17$other
imp.17$non.vigilant <- imp.17$foraging + imp.17$other

# Binomial regressions examining foraging/non-foraging in response to impala status within territorial herds.

################
##### NOTE #####
################   # we need to decide if these models will include a random effect or not 

# foraging model
bnom17.for <- glm(cbind(foraging,non.forage) ~ status, data = imp.17, family = binomial())

# Model summary
summary(bnom17.for)

# Using car Anova for Wald chisquare test to see if estimates significantly differ from 0. Type 3 if ranef, type 2 if not
car::Anova(bnom17.for, type = 2) 

# Using emmeans to assess pairwise comparisons via tukey posthoc test
emm.for.17 <- emmeans(bnom17.for,pairwise ~ status, type = "response") # We'll use this object for creating figures



# Vigilance model
bnom17.vig <- glm(cbind(vigilant,non.vigilant) ~ status, data = imp.17, family = binomial())

# Model summary
summary(bnom17.vig)

# Using car Anova for Wald chisquare test to see if estimates significantly differ from 0. Type 2 if ranef, type 2 if not
car::Anova(bnom17.vig, type = 3)

# Using emmeans to assess pairwise comparisons via tukey posthoc test
emm.vig.17 <- emmeans(bnom17.vig,pairwise ~ status, type = "response")

```
# Analyses for 2023, comparing differences in foraging and vigilance for males, females, and juveniles in territorial herds in WRF and KNP (and differences between 2 locations). Additionally will compare foraging and vigilance between males in territorial and bachelor herds in WRF and KNP (and differences between 2 locations)

```{r}
imp.23 <- read.csv("2023_individual.csv")

# Making new column of non-foraging and non-vigilance. needed for binomial regressions wherein we have success/failure format meaning success = targeted behavior and failure = not-targeted behavior.
imp.23$non.forage <- imp.23$vigilant + imp.23$other
imp.23$non.vigilant <- imp.23$foraging + imp.23$other


# Summarise by each of the groups.
sum23 <- imp.23 %>%
  group_by(location, status) %>%
  summarise(sum(total))

# Summarise by each of the groups.
sum23 <- imp.23 %>%
  group_by(location, status) %>%
  summarise(sum(total))

# Juveniles sample size is very low in KNP, they'll need removed from analyses
imp.23 <- imp.23 %>%
  filter(!(location == "KNP" & status == "juvenile"))

# Subset territorial data for modelling differences in behavior among individuals in territorial herds in WRF and KNP.
imp.23.ter <- imp.23 %>%
  filter(herd.type == "territorial") %>%
  droplevels()

# Binomial regression model examining foraging in territorial herds.
bnom23.ter.for <- glm(cbind(foraging,non.forage) ~ location*status, data = imp.23.ter, family = binomial())

# Model summary
summary(bnom23.ter.for)

# Using car Anova for Wald chisquare test to see if estimates significantly differ from 0. Type 3 due to presence of interactions
car::Anova(bnom23.ter.for, type = 2) 

# Using emmeans to gather probabilities and SE for each group. Will use for plotting
em.bn23.ter.for <- emmeans(bnom23.ter.for,  ~ location * status, type = "response")

# Contrasts for pairwise comparisons via Tukey tests
contrast(em.bn23.ter.for, method = "pairwise", by = "location")
contrast(em.bn23.ter.for, method = "pairwise", by = "status")



# Binomial regression model examining vigilance in territorial herds. 
bnom23.ter.vig <- glm(cbind(vigilant,non.vigilant) ~ location*status, data = imp.23.ter, family = binomial())

# Model summary
summary(bnom23.ter.vig)

# Using car Anova for Wald chisquare test to see if estimates significantly differ from 0. Type 3 due to presence of interactions
car::Anova(bnom23.ter.vig, type = 2)

# Making emmeans object for plotting and reporting
em.bn23.ter.vig <- emmeans(bnom23.ter.vig,  ~ location * status, type = "response")

# Contrasts for pairwise comparisons
contrast(em.bn23.ter.vig, method = "pairwise", by = "location")
contrast(em.bn23.ter.vig, method = "pairwise", by = "status")



# Compare foraging and vigilance between males in territorial and bachelor herds in WRF and KNP (and differences between 2 locations)
# Subset data for males 
imp.23.m <- imp.23 %>%
  filter(status == "b.male" | status == "t.male") %>%
  droplevels()

# Binomial regression
bnom23.bac.for <- glm(cbind(foraging,non.forage) ~ location*status, data = imp.23.m, family = binomial())

# Model summary
summary(bnom23.bac.for)

# Using car Anova for Wald chisquare test to see if estimates significantly differ from 0. Type 3 due to presence of interactions
car::Anova(bnom23.bac.for, type = 2) 

# Making emmeans object
em.bn23.bac.for <- emmeans(bnom23.bac.for,  ~ location * status, type = "response")

# Contrasts for pairwise comparisons
contrast(em.bn23.bac.for, method = "pairwise", by = "location")
contrast(em.bn23.bac.for, method = "pairwise", by = "status")



# Vigilance model
bnom23.bac.vig <- glm(cbind(vigilant,non.vigilant) ~ location*status, data = imp.23.m, family = binomial())

# Model summary
summary(bnom23.bac.vig)

# Using car Anova for Wald chisquare test to see if estimates significantly differ from 0. Type 3 due to presence of interactions
car::Anova(bnom23.bac.vig, type = 2) 

# Making emmeans object for plotting and reporting
em.bn23.bac.vig <- emmeans(bnom23.bac.vig,  ~ location * status, type = "response")

# Contrasts for pairwise comparisons
contrast(em.bn23.bac.vig, method = "pairwise", by = "location")
contrast(em.bn23.bac.vig, method = "pairwise", by = "status")
```
# Building plots

```{r}
library(ggplot2)
library(patchwork)
library(ggpattern)
# Reminder of emmeans summary names
emm.for.17 # 2017 foraging model territorial herds wits
emm.vig.17 # 2017 vigilance model territorial herds wits
em.bn23.ter.for # 2023 foraging model territorial herds WRF & KNP
em.bn23.ter.vig # 2023 vigilance model territorial herds WRF & KNP
em.bn23.bac.for # 2023 foraging model territorial males and bachelor males WRF & KNP 
em.bn23.bac.vig # 2023 vigilance model territorial males and bachelor males WRF & KNP

# Turning above summaries into dfs for plotting
foraging.17.em <- data.frame(emm.for.17$emmeans)
vigilant.17.em <- data.frame(emm.vig.17$emmeans)

foraging.23.ter.em <- data.frame(em.bn23.ter.for) %>%
  slice(-3)
vigilant.23.ter.em <- data.frame(em.bn23.ter.vig) %>%
  slice(-3)

foraging.23.bac.em <- data.frame(em.bn23.bac.for)
vigilant.23.bac.em <- data.frame(em.bn23.bac.vig)

# Setting theme for plots that I like
theme <- theme_classic()+
  theme(axis.title = element_text(size = 13), text = element_text(size = 11), axis.text.x=element_text(size=13, color = "black"), axis.text.y=element_text(size=11, color = "black"), legend.title=element_text(size = 12), legend.text=element_text(size=11), legend.position="right", axis.title.x = element_blank(), plot.title = element_text(size = 14, hjust = 0.5))

# Foraging plot
foraging.17.plot <- ggplot(data = foraging.17.em, aes(y = prob, x = status, fill = status)) + 
  geom_bar(position = "dodge", stat = "identity", alpha = .65, color = "black", linewidth = .75) +
  geom_errorbar(aes(ymin=(prob - SE), ymax = (prob + SE)), width = 0.3, linewidth = .75) + 
  theme +
    scale_fill_manual(values = c("#488650", "#C7C3C3", "#4C0C0B"), labels = c("Female", "Juvenile", "Territorial male"), name = "Impala type")+
  ylab("Probability") +
theme(axis.title.x = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1)) +
scale_x_discrete(limits= c( "female", "t.male", "juvenile"), labels=c("Female", "Territorial male", "Juvenile")) + 
  ggtitle("Foraging") + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) + 
  guides(fill = guide_legend(override.aes = list(linewidth = .25)))

# Vigilance plot
vigilance.17.plot <- ggplot(data = vigilant.17.em, aes(y = prob, x = status, fill = status)) + 
  geom_bar(position = "dodge", stat = "identity", alpha = .65, color = "black", linewidth = .75) +
  geom_errorbar(aes(ymin=(prob - SE), ymax = (prob + SE)), width = 0.3, linewidth = .75) + 
  theme +
    scale_fill_manual(values = c("#488650", "#C7C3C3", "#4C0C0B"), labels = c("Female", "Juvenile", "Territorial male"), name = "Impala type")+
  ylab("Probability") +
theme(axis.title.x = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1)) +
scale_x_discrete(limits= c( "female", "t.male", "juvenile"), labels=c("Female", "Territorial male", "Juvenile")) +
  ggtitle("Vigilance")  + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) + 
  guides(fill = guide_legend(override.aes = list(linewidth = .25)))

# Combining the 2 plots
plot.17 <- (foraging.17.plot | vigilance.17.plot) + plot_layout(guides = "collect")
ggsave("plot.17.fe.png", plot.17, width = 10, height = 6, dpi = 500)



##################
### 2023 plots ###
##################

# territorial comparisons in WRF AND KNP
foraging.23.ter.plot <- ggplot(data = foraging.23.ter.em, aes(y = prob, x = interaction(status, location), fill = status, pattern = location)) + 
  geom_bar_pattern(position = "dodge", stat = "identity", alpha = .65, color = "black", linewidth = .75, pattern_fill = "black", pattern_angle = 45, pattern_density = .1, pattern_spacing = .045, pattern_key_scale_factor = 0.6) + 
  scale_pattern_manual(values = c(KNP = "stripe", WRF = "none"), name = "Location") +
  geom_errorbar(aes(ymin=(prob - SE), ymax = (prob + SE)), width = 0.3, linewidth = .75) + 
  theme +
  scale_fill_manual(values = c("#488650", "#C7C3C3", "#4C0C0B"), labels = c("Female", "Juvenile", "Territorial male"), name = "Impala type")+
  ylab("Probability") +
theme(axis.title.x = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1)) +
scale_x_discrete(limits= c("female.WRF", "t.male.WRF", "juvenile.WRF", "female.KNP", "t.male.KNP"), labels=c("WRF Female", "WRF Territorial male", "WRF Juvenile", "KNP Female", "KNP Territorial male")) +
  ggtitle("Foraging") + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) + 
  guides(pattern = guide_legend(reverse = T), fill = guide_legend(override.aes = list(pattern = "none")))

# Vigilance
vigilance.23.ter.plot <- ggplot(data = vigilant.23.ter.em, aes(y = prob, x = interaction(status, location), fill = status, pattern = location)) + 
  geom_bar_pattern(position = "dodge", stat = "identity", alpha = .65, color = "black", linewidth = .75, pattern_fill = "black", pattern_angle = 45, pattern_density = .1, pattern_spacing = .045, pattern_key_scale_factor = 0.6) + 
  scale_pattern_manual(values = c(KNP = "stripe", WRF = "none"), name = "Location") +
  geom_errorbar(aes(ymin=(prob - SE), ymax = (prob + SE)), width = 0.3, linewidth = .75) + 
  theme +
  scale_fill_manual(values = c("#488650", "#C7C3C3", "#4C0C0B"), labels = c("Female", "Juvenile", "Territorial male"), name = "Impala type")+
  ylab("Probability") +
theme(axis.title.x = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1)) +
scale_x_discrete(limits= c("female.WRF", "t.male.WRF", "juvenile.WRF", "female.KNP", "t.male.KNP"), labels=c("WRF Female", "WRF Territorial male", "WRF Juvenile", "KNP Female", "KNP Territorial male")) +
  ggtitle("Vigilance") +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) + 
  guides(pattern = guide_legend(reverse = T), fill = guide_legend(override.aes = list(pattern = "none")))
                                 
# Combining the 2 plots
terr.plot.23 <- (foraging.23.ter.plot | vigilance.23.ter.plot) + plot_layout(guides = "collect")
ggsave("terr.plot.23.fe.png", terr.plot.23, width = 10, height = 6, dpi = 500)


# Male comparisons in WRF and KNP
foraging.23.male.plot <- ggplot(data = foraging.23.bac.em, aes(y = prob, x = interaction(status, location), fill = status, pattern = location)) + 
  geom_bar_pattern(position = "dodge", stat = "identity", alpha = .65, color = "black", linewidth = .75, pattern_fill = "black", pattern_angle = 45, pattern_density = .1, pattern_spacing = .045, pattern_key_scale_factor = 0.6) + 
  scale_pattern_manual(values = c(WRF = "none", KNP = "stripe"), name = "Location") +
  geom_errorbar(aes(ymin=(prob - SE), ymax = (prob + SE)), width = 0.3, linewidth = .75) + 
  theme +
  scale_fill_manual(values = c("#AB6E24", "#4C0C0B"), labels = c("Bachelor male", "Territorial male"), name = "Impala type") +
  ylab("Probability") +
theme(axis.title.x = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1)) +
scale_x_discrete(limits= c("t.male.WRF", "b.male.WRF", "t.male.KNP", "b.male.KNP"), labels=c("WRF Territorial male", "WRF Bachelor male", "KNP Territorial male", "KNP Bachelor male")) +
  ggtitle("Foraging") +  
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) + 
  guides(pattern = guide_legend(reverse = T), fill = guide_legend(override.aes = list(pattern = "none")))

# Vigilance
vigilance.23.male.plot <- ggplot(data = vigilant.23.bac.em, aes(y = prob, x = interaction(status, location), fill = status, pattern = location)) + 
  geom_bar_pattern(position = "dodge", stat = "identity", alpha = .65, color = "black", linewidth = .75, pattern_fill = "black", pattern_angle = 45, pattern_density = .1, pattern_spacing = .045, pattern_key_scale_factor = 0.6) + 
  scale_pattern_manual(values = c(WRF = "none", KNP = "stripe"), name = "Location") +
  geom_errorbar(aes(ymin=(prob - SE), ymax = (prob + SE)), width = 0.3, linewidth = .75) + 
  theme +
  scale_fill_manual(values = c("#AB6E24", "#4C0C0B"), labels = c("Bachelor male", "Territorial male"), name = "Impala type") +
  ylab("Probability") +
theme(axis.title.x = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1)) +
scale_x_discrete(limits= c("t.male.WRF", "b.male.WRF", "t.male.KNP", "b.male.KNP"), labels=c("WRF Territorial male", "WRF Bachelor male", "KNP Territorial male", "KNP Bachelor male")) +
  ggtitle("Vigilance") + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) + 
  guides(pattern = guide_legend(reverse = T), fill = guide_legend(override.aes = list(pattern = "none")))

# Combining the 2 plots
male.plot.23 <- (foraging.23.male.plot | vigilance.23.male.plot) + plot_layout(guides = "collect")
ggsave("male.plot.23.fe.png", male.plot.23, width = 10, height = 6, dpi = 500)

```
